---
title: "Intro to data wrangling and visualization with the tidyverse"
subtitle: "Workshop 1, Module 3"
format: qmdrslides-revealjs
editor: source
execute:
  echo: true
  message: false
  warning: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(countdown)
library(here)

table_01 <- read_csv(here("datasets/instructional_dataset/01_participant_metadata_UKZN_workshop_2023.csv"))
table_02 <- read_csv(here("datasets/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))
```

## Goals for this module

By the end of this module you can:

- look at a table and say what one row represents
- pick rows with `filter()` and columns with `select()`
- derive new columns with `mutate()`
- sort rows with `arrange()`
- count categories with `count()`
- recognise verbs that keep what a row represents the same vs verbs that change it
- read and write a basic ggplot: data, mapping, geom

## The pipe `|>`

`|>` is pronounced "and then".

`table_02 |> head()` reads as "take `table_02`, and then call `head()` on it."

R 4.1 added `|>` to base R. We use it everywhere in this module.

One quirk: the right-hand side needs parentheses. Write `head()`, not `head`.

## The pipe in action {auto-animate=true}

```{r}
head(table_02)
```

## The pipe in action {auto-animate=true}

```{r}
table_02 |> head()
```

## The pipe in action {auto-animate=true}

```{r}
table_02 |>
  head() |>
  nrow()
```

## What does a row represent?

```{r}
table_02
```

132 rows. Each row is one participant at one visit (44 participants × 3 visits).

The whole wrangling half asks the same question: as we work on this table, does a row keep meaning the same?

## `filter()` picks rows {auto-animate=true}

![](images/filter.svg)

## `filter()` picks rows {auto-animate=true}

```{r}
table_02 |>
  filter(arm == "placebo")
```

Same kind of row (a participant-visit). Fewer of them.

## `filter()` with logical operators

```{r}
table_02 |> filter(ph < 4)
```

```{r}
table_02 |> filter(arm == "placebo", time_point == "early")
```

Also useful: `&`, `|`, `%in%`, `is.na()`.

## `select()` picks columns {auto-animate=true}

![](images/select.svg)

## `select()` picks columns {auto-animate=true}

```{r}
table_02 |>
  select(pid, arm, ph, nugent_score)
```

Same rows, fewer columns. Use `-column` to drop a column. Use `new = old` to rename inside `select()`.

## `mutate()` adds derived columns {auto-animate=true}

![](images/mutate.svg)

## `mutate()` adds derived columns {auto-animate=true}

```{r}
table_02 |>
  mutate(crp_blood_ugul = crp_blood / 1000)
```

Same rows, one more column.

## `mutate()` with `if_else()`

```{r}
table_02 |>
  mutate(high_ph = if_else(ph >= 4.5, "yes", "no")) |>
  select(pid, time_point, ph, high_ph) |>
  head()
```

`if_else(condition, value_if_true, value_if_false)`. The new column has one value per row.

## `arrange()` sorts rows {auto-animate=true}

![](images/arrange.svg)

## `arrange()` sorts rows {auto-animate=true}

```{r}
table_02 |> arrange(ph)
```

```{r}
table_02 |> arrange(desc(ph))
```

Same rows, reordered. The table's meaning is unchanged.

## A growing pipeline {auto-animate=true}

```{r}
table_02 |>
  filter(time_point == "early")
```

## A growing pipeline {auto-animate=true}

```{r}
table_02 |>
  filter(time_point == "early") |>
  select(pid, arm, ph, nugent_score, crp_blood)
```

## A growing pipeline {auto-animate=true}

```{r}
table_02 |>
  filter(time_point == "early") |>
  select(pid, arm, ph, nugent_score, crp_blood) |>
  mutate(high_ph = if_else(ph >= 4.5, "yes", "no"))
```

## A growing pipeline {auto-animate=true}

```{r}
table_02 |>
  filter(time_point == "early") |>
  select(pid, arm, ph, nugent_score, crp_blood) |>
  mutate(high_ph = if_else(ph >= 4.5, "yes", "no")) |>
  arrange(desc(ph))
```

## `count()` — the bridge to Module 4

```{r}
table_01 |> count(smoker)
```

`table_01` has 44 rows, one per participant. `count(smoker)` returns 2 rows.

A row is no longer a participant. It is a smoker-category.

This is the first verb that changes what a row represents. Module 4 is mostly verbs like this.

## `count()` on multiple columns

```{r}
table_01 |> count(arm, smoker)
```

Each row is one arm × smoker combination.

## Exercise 1 {.your-turn}

Open `notebooks/03-wrangling.qmd` in your Codespace.

Group A: agentic. Group B: manual.

```{r}
countdown::countdown(30)
```
