<!-- Included partial: rendering is controlled by the parent index.qmd. -->

## Worksheet: build your first R notebook

This is the hands-on half of the session. Over the next hour you will build up
**one** Quarto notebook from an empty file, running code and writing a little
prose as you go. By the end you will have a small report you wrote yourself.

Open `notebooks/02-intro-to-r.qmd` in your Codespace and keep this page open
next to it. The notebook already has a heading and an empty code chunk for each
step below -- you fill them in here.

The whole point of this worksheet is one habit, repeated until it feels normal:

> **write a chunk → run it → write a sentence about what you saw → render.**

### How to read this worksheet

Every instruction below is one of four kinds. Each has its own coloured box, so
you always know what you are being asked to do. Here is one of each:

::: {.callout-note title="▶️ Run this"}
Copy the chunk into your notebook and click **Run Cell** above it. The output
appears right under the chunk.
:::

::: {.callout-tip title="✍️ Your turn"}
Write the code or the prose yourself. There is an empty chunk waiting for it in
your notebook.
:::

::: {.callout-important title="🤖 Ask Pi"}
Open a terminal, type `pi`, and ask Pi for help. Read what it gives you before
you use it.
:::

::: {.callout-caution title="🔄 Render and look"}
Click **Preview** at the top right of the notebook. Read your work as a webpage,
the way someone else would.
:::

When you see one of these four boxes, do that one thing. That is all there is to
it.

## Part 1 — Run a chunk, write a sentence, render

*About 10 minutes.* The goal here is just the loop. The code is simple on
purpose.

Every notebook starts by loading the packages it needs. You already have a
setup chunk at the top of `02-intro-to-r.qmd`:

::: {.callout-note title="▶️ Run this"}
This is the setup chunk at the top of your notebook. Run it first, every time
you open the notebook.

```{r}
#| eval: false
library(tidyverse)
library(here)
```
:::

Now some arithmetic. R is a calculator before it is anything else.

::: {.callout-note title="▶️ Run this"}
Put this in the **Part 1** chunk and run it. The answers appear inline.

```{r}
#| eval: false
3 + 5 + 10
10 * (5 + 1)
2 ** 10
```
:::

::: {.callout-tip title="✍️ Your turn"}
Above the chunk, write one plain sentence in your own words, for example: *"R
treats `**` as 'to the power of', so `2 ** 10` is 2 multiplied by itself ten
times."* This is the prose half of a notebook -- code surrounded by your own
explanation.
:::

::: {.callout-note title="▶️ Run this"}
Add these two lines below the others and run again. They look almost the same.

```{r}
#| eval: false
64 / 2 + 2
64 / (2 + 2)
```
:::

::: {.callout-tip title="✍️ Your turn"}
Write a sentence explaining why the two lines give different answers. (Hint:
what do the brackets change?)
:::

::: {.callout-caution title="🔄 Render and look"}
Click **Preview**. You should see your sentences and your code and its output,
laid out as a clean page. That is the loop. You will do it again four more
times.
:::

## Part 2 — Read the data and look at it

*About 13 minutes.* From here on you work with the real workshop dataset: the
clinical measurements table (`02_visit_clinical_measurements`) you saw in the
slides. One row is one participant at one visit.

::: {.callout-tip title="✍️ Your turn"}
In the **Part 2** chunk, read the file into a tibble called `table_02`, then
print it. Type it yourself:

```{r}
#| eval: false
table_02 <- read_csv(here(
  "data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))

table_02
```

`<-` saves the data under the name `table_02`. After this line, `table_02` *is*
the data.
:::

::: {.callout-tip title="✍️ Your turn"}
Read the printout. Write a sentence recording how many rows and how many columns
it has, and name two columns that look interesting to you.
:::

::: {.callout-note title="▶️ Run this"}
`glimpse()` is a friendlier way to see every column at once. Add it and run.

```{r}
#| eval: false
glimpse(table_02)
```
:::

::: {.callout-caution title="🔄 Render and look"}
Render again. Notice that a printed tibble and a `glimpse()` look different in
the report -- both are useful for getting your bearings in a new dataset.
:::

## Part 3 — Pull out columns and summarise them

*About 13 minutes.* A tibble is a stack of columns. Each column is a **vector**:
a sequence of values, all the same type. You can pull one out and do maths on it.

::: {.callout-tip title="✍️ Your turn"}
In the **Part 3** chunk, pull out the `nugent_score` column with `$` and run it.

```{r}
#| eval: false
table_02$nugent_score
```

That row of numbers is the vector. The Nugent score runs 0--10; 7 or above
suggests bacterial vaginosis.
:::

::: {.callout-tip title="✍️ Your turn"}
Now summarise a column with a function. `mean()` takes a vector and gives back
one number. Add this and run:

```{r}
#| eval: false
mean(table_02$crp_blood, na.rm = TRUE)
```

`na.rm = TRUE` tells `mean()` to skip missing values. Try removing it and
running again -- what happens, and why?
:::

::: {.callout-tip title="✍️ Your turn"}
Save the result under a name so you can reuse it, then print the name:

```{r}
#| eval: false
crp_mean <- mean(table_02$crp_blood, na.rm = TRUE)
crp_mean
```

Write a sentence reporting the mean blood CRP and what `na.rm = TRUE` did.
:::

::: {.callout-caution title="🔄 Render and look"}
Render. You now have prose, a vector, a summary, and a saved value in one
report.
:::

## Part 4 — Build something with Pi

*About 14 minutes.* Pi is the coding agent from Module 1. It runs in the
terminal and can read the files in your project. Here you use it the way you
should all week: as a fast helper whose work you always check.

::: {.callout-important title="🤖 Ask Pi"}
Open a terminal (**Terminal → New Terminal**), type `pi`, and ask it something
specific, naming the file and the columns:

> Read `data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv`
> and write an R chunk that shows the mean `nugent_score` for each `arm`.
:::

::: {.callout-note title="▶️ Run this"}
Paste Pi's chunk into the **Part 4** chunk and run it. Did it work first try? If
it errored, read the error, paste it back to Pi, and try its fix.
:::

::: {.callout-important title="🤖 Ask Pi"}
Push back on the code, even if it ran. Ask:

> Explain that chunk to me line by line. Why did you use those functions?

Notice when Pi gives a clear, correct explanation and when it sounds confident
but vague.
:::

::: {.callout-tip title="✍️ Your turn"}
Write a sentence on two things: did Pi's output actually answer what you asked,
and did the column names it used really exist in the data? (Pi sometimes invents
column names -- always check.)
:::

::: {.callout-caution title="🔄 Render and look"}
Render. Part of the skill is making Pi's contribution fit cleanly into a report
that reads as yours.
:::

## Capstone — make it your own

*About 14 minutes.* No step-by-step this time. Build a short piece of analysis
about something in this dataset you are actually curious about. The columns
`nugent_score`, `crp_blood`, `ph`, `arm`, and `time_point` are all fair game.

::: {.callout-tip title="✍️ Your turn"}
In the **Capstone** chunk, pick one numeric column and compute a few summaries
of it -- some of `min()`, `max()`, `median()`, `mean()`, `sd()`. Run as you go.
:::

::: {.callout-important title="🤖 Ask Pi"}
Use Pi **once** to push your analysis a step further. For example: *"How many
visits have a `nugent_score` of 7 or higher?"* or *"Show the mean `ph` at each
`time_point`."* Run its answer and check it.
:::

::: {.callout-tip title="✍️ Your turn"}
Write a short paragraph -- three or four sentences -- saying what you looked at
and what you found. Connect it to the biology where you can (for instance, a
Nugent score of 7+ is the bacterial-vaginosis range).
:::

::: {.callout-caution title="🔄 Render and look"}
Render one last time. This is your notebook. Read it top to bottom as a report
and fix anything that does not flow.
:::

## Before you go

::: {.callout-important title="🤖 Ask Pi"}
Pick one step today that confused you. Ask Pi to explain it, then push back:
*"why did you use X instead of Y?"* Notice when Pi sounds confident versus when
it is guessing. Using Pi as a study partner rather than an answer machine is the
habit that will keep you learning all week.
:::

::: {.callout-note collapse="true" title="📖 Cheat-sheet: R you might want while you work"}
You do not need to read this to finish the worksheet. It is here for reference
when you want a reminder.

**Arithmetic**

```r
+  -  *  /        add, subtract, multiply, divide
^  or  **         power: 2 ** 3 is 8
x %% y            remainder: 5 %% 2 is 1
x %/% y           whole-number division: 5 %/% 2 is 2
```

**Comparisons** (give `TRUE` or `FALSE`)

```r
<   <=   >   >=   less / greater than (or equal)
==               equal to
!=               not equal to
x & y            x AND y
x | y            x OR y
```

**Summarising a column**

```r
mean(x, na.rm = TRUE)   average, skipping missing values
median(x)               middle value
sd(x)                   standard deviation
min(x)   max(x)         smallest / largest
range(x)                min and max together
sum(x)                  total
length(x)               how many values
```

**Looking at a tibble**

```r
table_02              print it
glimpse(table_02)     every column, one per line
head(table_02)        first few rows
nrow(table_02)        number of rows
ncol(table_02)        number of columns
names(table_02)       the column names
table_02$column_name  pull out one column as a vector
```

**Getting help**

```r
?mean        open the help page for a function
?read_csv
```

You can also ask Pi to explain a function -- then check what it says against the
help page.
:::

::: {.callout-tip collapse="true" title="Going further (optional)"}
If you finish early and want more, try any of these in your notebook:

1. Make your own small tibble from scratch with `tibble()` -- a few columns of
   your own data -- and summarise one of them.
2. Compare a measure between the two arms: is the mean `crp_blood` different for
   `placebo` versus `treatment`? (Ask Pi how to do this with `group_by()` and
   `summarise()`, then check the result.)
3. Pull out just the baseline visits and summarise `ph` for them. How would you
   describe, in one sentence, what "baseline" means here?
:::
