---
title: Modeling and Uncertainty
date: 2026-05-20
bibliography: references.bib
format:
  html:
    embed-resources: true
    execute:
      message: false
      warning: false
---

```{r}
#| label: load-libraries
library(MASS)
library(lme4)
library(lmerTest)
library(patchwork)
library(modelsummary)
library(survival)
library(tidyverse)
source("helpers.R")
theme_set(theme_classic())
set.seed(202606)
```

## Note on Data Generation

The data used in this document are read directly from online sources. The original data generation code has been externalized to `@helpers.qmd` to keep the main document focused on analysis and interpretation.

## Learning Outcomes

1. Explain models as a description of how we think the data were generated, and
distinguish between the associated "large" and "small" worlds.

1. Differentiate between the members of the biostatistical model zoo and explain
the types of outcomes and questions to which they are well-suited.

1. Use R to implement these models, using formula syntax to specify the
relationship between covariates and outcomes, and interpret the estimated model
parameters in the original problem context.

1. Diagnose model limitations using visualization and simulation.

1. Describe how to improve model fit by modifying assumptions (adding
covariates, random effects, interactions).

## Part 1: What is a Model?

1. Below are a few views of IL-1a concentration from the workshop dataset.
Notice the variation across people and a possible treatment effect at the
post-treatment visit.

   ```{r}
#| label: load-data-il1
ids <- read_csv("data/00_sample_ids_UKZN_workshop_2023.csv")
il1 <- read_csv("data/03_elisa_cytokines_UKZN_workshop_2023.csv") |>
    filter(cytokine == "IL-1a") |>
    left_join(ids)
head(il1)
   ```

   ```{r}
#| label: plot-il1
#| fig-width: 6
#| fig-height: 4
p1 <- ggplot(il1) +
    geom_line(aes(time_point, log(conc), group = pid, col = arm)) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_discrete(expand = c(0, 0))

p2 <- ggplot(il1) +
    geom_histogram(aes(log(conc), fill = arm), bins = 10, position = "identity", alpha = 0.6) +
    facet_wrap(~ time_point)

p1 / p2
   ```

1. Imagine running the study again, keeping the recruitment and experimental
protocols as similar as possible. Make two lists,

   - Aspects of these plots that you think will continue to be true in the new
  experiment.
   - Aspects of these plots that you think will be different.

   Take 5 minutes to prepare your description.

1. A statistical model is a simplified description of how we think the data were
generated (the "data story"). Its fitted parameter lets us answer substantive
scientific questions, like how a treatment affects outcomes.

     [picture of the DGP -> experiment]

1. **Large vs. small worlds**. We should distinguish the "large world" where the
data were gathered from the "small world" of the statistical model. The small
world makes its assumptions precise, allows hypothetical interventions, and
quantifies uncertainty. But there will always be a gap between the two worlds,
and that is itself a source of uncertainty not captured by the model.

1. **Modeling Checklist.** Organize the work of modeling into these steps.
steps.

   - _Biological Question._ A good model should clarify our thinking about a
   real phenomenon and not be made just for the sake of modeling.
   - _Plot the Data._ What does the outcome look like? Do you
   already notice any relationships with other variables?
   - _Working Assumptions._ Make a small-world story for how the data were
   generated. This is the initial statistical model.
   - _Model Checks._ Use simulation and diagnostics to check whether the model
   captures patterns we see in the real data.
   - _Revised Models._ Revise the working assumptions until the model provides
   a good match. This model can now be used to answer the original biological
   question.

## Part 2: A Tour through the Model Zoo

Each type of statistical model is designed for different kind of data or
biological question.  Knowing all the "species" of model will let you adapt to
whatever data or questions you encounter.

### Bernoulli and Binomial Models

1. Bernoulli and binomial models apply when outcomes fall into two groups. The
bernoulli models individual outcomes while the binomial applies to their totals.

   - Infected vs. not infected
   - Responder vs. non-responder
   - Alive vs. dead

1. The data story is that each person's characteristics shift the relative
probabilities of the two outcomes. In this case, more social contact, higher
probability of infection. The bernoulli applies when we observe person-level
data.

   ```{r}
#| label: load-bernoulli-data
infections <- read_csv("https://huggingface.co/datasets/krisrs1128/durban_workshop_data/resolve/main/infections.csv")
   ```

   ```{r}
#| label: show-bernoulli-data
infections
   ```

   ```{r}
#| label: plot-bernoulli
#| fig-width: 5.5
#| fig-height: 3.5
ggplot(infections) +
    geom_jitter(aes(social_contact, as.numeric(infected) - 1, col = infected), height = 0.05) +
    geom_line(aes(social_contact, prob_infection), linewidth = 1, col = "#232323") +
    scale_x_continuous(expand = c(0, 0)) +
    labs(
        y = "Probability of Infection",
        x = "Level of Social Contact",
        col = "Infection Status"
    )
   ```

1. The binomial model applies when we see totals within categories, but
not person-level outcomes.

   ```{r}
#| label: load-binomial-data
infections_props <- read_csv("https://huggingface.co/datasets/krisrs1128/durban_workshop_data/resolve/main/infections_props.csv")
   ```

   ```{r}
#| label: show-binomial-data
infections_props
   ```

   ```{r}
#| label: plot-binomial
#| fig-width: 5.5
#| fig-height: 2.5
infections_props |>
    select(-total) |>
    pivot_longer(-social_level) |>
    ggplot() +
    geom_col(aes(social_level, value, fill = name), position = "dodge") +
    scale_y_continuous(expand = c(0, 0, 0.1, 0)) +
    labs(
        x = "Level of Social Contact",
        y = "Count",
        fill = "Infection Status"
    )
   ```

1. To fit these models, we can use the `glm` function. This function has three
arguments,

   - `formula`: A formula  like `y ~ x` specifies the relationship between the
   model outcome `y` and contextual variables `x`, which we will call
   "covariates."
   - `family`: The type of model to use for the given dataset. This depends on
   the data type for the outcome (binary, count, etc.) and it differs across the
   examples in this section.
   - `data`: A `data.frame` object whose columns give the response and predictor
   variables to use in the model. Any terms appearing in the formula must be
   column names in this `data.frame`.

   For example, for a bernoulli model to have probability of infection depend on
   `social_contact`, we can use the command below. Even though we set
   `family = binomial()`, the fact that `infected` is a 0/1 variable means that internally
   the command uses a bernoulli model^[The technical reason is that the
   bernoulli distribution is a special case of the binomial when the total count
   parameter is set to 1.].

   ```{r}
#| label: bernoulli-glm
fit <- glm(infected ~ social_contact, family = binomial(), data = infections)
modelsummary(fit, shape = term ~ statistic, gof_map = NA)
   ```

   The estimate column describes how infection probability changes across levels
   of social contact. As given, they are not easily interpretable^[Because they
   are given on the log-odds scale. The model estimation algorithm works on this
   scale because it isn't constrained to $\left[0, 1\right]$, and so
   general-purpose optimization methods can be used.]. The `plogis` function
   converts them to the probability scale.

   ```{r}
#| label: plogis-eval
plogis(-2.37) # when social contact = 0
plogis(-2.37 + 0.5 * 3.09) # when social contact = 0.5
plogis(-2.37 + 1 * 3.09) # when social contact = 1
   ```

   This bernoulli model is often called _logistic_ regression, after the inverse
   _logistic_ function (`plogis`) that converts model estimates to the
   probability scale.

   The `infection` data were simulated with a true `social_contact` effect of 3.
   The estimate 2.37 is close, but even with 100 samples, slightly off.
   Nonetheless, it lies within one standard error (0.79) of the truth -- the
   model gives both an estimate of the effect and a measure of its uncertainty.

1. The same function fits the binomial model, but the formula's left-hand side
needs to account for group-level counts, not person-level outcomes. Also, the
`weights` argument is added to reflect the different totals in each category.

   ```{r}
#| label: binomial-glm
fit <- glm(infected / total  ~ social_level, family = binomial(), weights = total, data = infections_props)
modelsummary(fit, shape = term ~ statistic, gof_map = NA)
   ```

### Poisson and Negative Binomial Models

1. Poisson and negative binomial models apply to count data. The poisson arises
whenever we count events spread uniformly across space and time, for example,
the number of raindrops landing on a window during a minute of rain.  Many
biological counts have this flavor.

    - Transcripts counts in a cell
    - Mutation count along a segment of the genome
    - Infection count within a county

   _Exercise: The last example is again about infections, which we saw in the bernoulli and binomial examples. What did we know there we don't know from infection counts alone?_

1. In the data story, the typical count depends on sample characteristics. For
example, more infections in counties closer to the outbreak epicenter.  This
mean parameter is called the poisson rate.

   ```{r}
#| label: load-poisson-data
outbreak <- read_csv("https://huggingface.co/datasets/krisrs1128/durban_workshop_data/resolve/main/outbreak.csv")
   ```

   ```{r}
#| label: poisson-view
outbreak
   ```

   ```{r}
#| label: poisson-plot
#| fig-width: 5.5
#| fig-height: 3
ggplot(outbreak, aes(epicenter_distance)) +
    geom_line(aes(y = mean), col = "#414ca1", linewidth = 1) +
    geom_point(aes(y = infected)) +
    labs(
        x = "Epicenter Distance",
        y = "Infection Count"
    )
   ```

1. As before, `glm` applies, but we set the `family` argument to `poisson()`.

   ```{r}
fit <- glm(infected ~ epicenter_distance, family = poisson(), data = outbreak)
modelsummary(fit, shape = term ~ statistic, gof_map = NA)
   ```

   The model output is reported on a log scale, so it should be interpreted
   after exponentiation. The fitted mean at distance 0 is `exp(2.5) = 12.8`,
   matching the plot. Similarly, larger distances are interpreted after plugging
   into a linear formula and exponentiating. For example, at distance 0.5, the
   mean is `exp(2.5 - .5 * 4.3) = 1.4`.

   _Exercise: What is the level of uncertainty in these estimates? How would you interpret 0.08 in the `S.E.` column? What about 0.29?_

1. Often unmeasured factors influence the rates, making the data more
heterogeneous than the poisson can handle. For example, infection might depend
not only distance to the epicenter but unknown, county-specific population
densities.  The negative binomial distribution addresses this -- it arises as a
mixture of poissons with similar but slightly different rates. In the data
simulated below, the counts at a given distance are more spread out even though
the blue mean line is unchanged.

   ```{r}
#| label: load-negbin-data
outbreak_nb <- read_csv("https://huggingface.co/datasets/krisrs1128/durban_workshop_data/resolve/main/outbreak_nb.csv")
   ```

   ```{r}
#| label: negbin-view
head(outbreak_nb)
   ```

   ```{r}
#| label: negbin_plot
#| fig-width: 5.5
#| fig-height: 3
ggplot(outbreak_nb, aes(epicenter_distance)) +
    geom_line(aes(y = mean), col = "#414ca1", linewidth = 1) +
    geom_point(aes(y = infected)) +
    labs(
        x = "Epicenter Distance",
        y = "Infection Count"
    )
   ```

1. `glm` doesn't provide a negative binomial model, but `glm.nb` from the `MASS`
package does, with the same syntax as `glm`.

   ```{r}
#| label: negbin_fit
fit <- glm.nb(infected ~ epicenter_distance, data = outbreak_nb)
modelsummary(fit, shape = term ~ statistic, gof_map = NA)
   ```

### Linear Regression Models

1. When the response is continuous, linear regression applies. For example, it
can be used for,

   - Antibody titre
   - CRP concentration

   The model assumes symmetric, random error around a mean value that depends on
   the inputs. For example, antibody titre might vary as a linear function of
   age.

   ```{r}
#| label: load-lm-data
antibody <- read_csv("https://huggingface.co/datasets/krisrs1128/durban_workshop_data/resolve/main/antibody.csv")
   ```

   ```{r}
#| label: antibody-view
head(antibody)
   ```

   ```{r}
#| label: plot-lm
#| fig-width: 5.5
#| fig-height: 3
ggplot(antibody) +
    geom_line(aes(age, mean), col = "#d53636", linewidth = 1.5) +
    geom_point(aes(age, antibody)) +
    labs(x = "Age", y = "Antibody Titre")
   ```

1. The `lm` function fits this model, with the same inputs as `glm`, but no
`family`. The estimates give predicted values at age $x$ $19.6 + 0.058 x$. For
example, at age 10, the prediction is $19.6 + 0.58 \approx 20.18$.

   ```{r}
#| label: lm-simple
fit <- lm(antibody ~ age, antibody)
modelsummary(fit, shape = term ~ statistic, gof_map = NA)
   ```

1. Linear regression assumes a Gaussian distributed error term.  This
distribution appears across statistics because of the central limit theorem: the
average of many small independent effects tends a Gaussian distribution.  In
this example, we're implicitly assuming many small, independent factors besides
age that perturb the observed antibody measurement.

### Survival Models

1. Survival models describe whether and when an event occurs. For example,

   - Time until heart attack
   - Time until loan default
   - Time until HIV acquisition

   If the event doesn't happen over the duration of the study, the study
   participant's response is called "censored."

1. Unlike the other models, where the outcome was a single number, the survival
analysis outcomes two parts.

   - The `time` the event was observed. Or, if the event was never observed, the
   time the study ended.
   - Whether the event occurred (`status = 1`) or not (`status = 0`) before the
   study end.

   For example, imagine modeling HIV acquisition as a function of an drug
   treatment, tracking participants for two years. Some have an event during the
   study period (`status = 1`, triangles) while the rest are censored.

   ```{r}
#| label: load-survival-data
surv_data <- read_csv("https://huggingface.co/datasets/krisrs1128/durban_workshop_data/resolve/main/surv_data.csv")
   ```

   ```{r}
#| label: survival-data
head(surv_data)
   ```

   ```{r}
#| label: survival-plot
#| fig-width: 6
#| fig-height: 5
surv_data |>
    mutate(id = row_number()) |>
    ggplot(aes(time, reorder(id, time), color = treatment, shape = factor(status))) +
     geom_point(size = 4) +
     labs(
       x = "Time",
       y = "Participant",
       color = "Treatment",
       shape = "Event?"
     ) +
     theme(panel.grid.major.y = element_line(linewidth = 0.1))
   ```

1. The most popular form of survival model is the Cox proportional hazards
model. During any time interval, anyone who hasn't yet had an event is at risk
for it. Some covariates can change the relative risk ("hazard") of the event.
Because a covariate shifts the hazard by the same factor at all times, the model
is called a _proportional_ hazards model.

   ```{r}
#| label: surv-fit
fit <- coxph(Surv(time, status) ~ treatment, data = surv_data)
modelsummary(fit, shape = term ~ statistic, gof_map = NA)
   ```

   The model output shows how each covariate rescales the hazard.

   [picture of hazard functions]

   _Exercise: Can you think of an example where the proportional hazards assumption does not hold? Hint: What if a drug was not effective immediately after treatment start but required several weeks to reach its full effect?_

### Variations: Multiple Predictors, Mixed Effects, and Interactions

1. **Multiple Predictors.** The examples above had a single covariate.  We can
add more covariates with `+ variable` on the right-hand side of the formula.
Each variables' estimate is the effect of that variable with all others held
fixed.

   For example, if we wanted to model antibody titre as a function of both age
   and blood pressure, we could use the formula `antibody ~ age +
   blood_pressure`. The interpretation of the estimate below is that, fixing
   blood pressure level, the average antibody level increases by 0.058 units for
   every year increase in age.

   ```{r}
fit <- lm(antibody ~ age + blood_pressure, data = antibody)
modelsummary(fit, shape = term ~ statistic, gof_map = NA)
   ```

1. **Mixed Effects.** One common situation is that baseline effects (overall
probabilities in bernoulli model, rates in the poisson, intercepts in the linear
regression, ...) vary across contexts. For example, infection probability might
always increase with social contact, but the baseline rate may vary from county
to county, due to unmeasured factors like population density and mobility.

   ```{r}
#| label: load-mixed-effect-data
mixed_effect_data <- read_csv("https://huggingface.co/datasets/krisrs1128/durban_workshop_data/resolve/main/mixed_effect_data.csv")
   ```

   ```{r}
#| label: show-mixed-effect-data
head(mixed_effect_data)
   ```

   ```{r}
#| label: plot-mixed-effect-data
ggplot(mixed_effect_data) +
    geom_jitter(aes(social_contact, as.numeric(infected) - 1, col = infected), height = 0.05) +
    geom_line(aes(social_contact, prob_infection), linewidth = 1, col = "#232323") +
    facet_wrap(~ reorder(county, prob_infection), nrow = 2) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_continuous(expand = c(0, 0)) +
    labs(y = "Probability of Infection")
   ```

1. One solution is to include county as an ordinary covariate.  But with many
counties this becomes unwieldy -- we would have a separate estimate for each.  A
better approach is to use a _random effect_ (also called a _hierarchical
model_), where the county-level effects are imagined to be drawn from a larger
source of randomness. Rather than estimate each county's effect separately, we
estimate the overall variation across counties. Statistically, this is an easier
task, and it yields higher power.  Because they include both fixed and random
effects, these models are called mixed effects models.

1. To implement this, we replace `glm` with `glmer` and add `(1 | county)` to
the formula, indicating that county is a random rather than fixed effect.

   ```{r}
#| label: random-county-glmer
fit <- glmer(infected ~ social_contact + (1 | county), family = binomial(), data = mixed_effect_data)
modelsummary(fit, shape = term ~ statistic, gof_map = NA)
   ```

   To interpret the result, look at `SD`row.  The 0.708 means that estimates
   vary by roughly this much across counties, at any level of `social_contact`.
   Even at social contact 0, the infection probabilities below would be typical.

   ```{r}
#| label: range-mixed-effect
plogis(-2.28 - 0.43) # lower range across counties, social contact = 0
plogis(-2.28 + 0.43) # upper range across counties, social contact = 0
   ```

1. **Interactions.** In some data, the effect of a variable on a response might
depend on other context. For example, the age effect might differ between
healthy and sick people.

   ```{r}
#| label: load-interaction-data
interaction_data <- read_csv("https://huggingface.co/datasets/krisrs1128/durban_workshop_data/resolve/main/interaction_data.csv")
   ```
   ```{r}
#| label: show-interaction-data
head(interaction_data)
   ```

   ```{r}
#| label: plot-interaction
#| fig-width: 5
#| fig-height: 3.5
ggplot(interaction_data) +
    geom_line(aes(age, mean, col = group), linewidth = 1.5) +
    geom_point(aes(age, antibody, col = group)) +
    labs(x = "Age", y = "Titre")
   ```

1. This type of pattern can be modeled using an interaction term.  The R formula
syntax has the form `y ~ x1 + x1 : x2`, meaning that the `x1` effect depends on
the level of `x2`.

   ```{r}
#| label: lm-interaction
fit <- lm(antibody ~ age + age : group, data = interaction_data)
modelsummary(fit, shape = term ~ statistic, gof_map = NA)
   ```

   In the healthy group, antibody titre rises by 0.059 units per year of age,
   while in the disease group, it changes by 0.0259 + 0.002 units (adding the
   two rows).

1. Although we introduced these variations with specific bernoulli and linear
regression models, they apply to any model above. They come in handy when
revising a model's working assumptions to improve its fit.

## Part 3: Case Studies

### Longitudinal Cytokines Analysis

1. **Biological Question.** In our study of interest, does treatment with the
anti-inflammatory drug change cytokine concentrations post-treatment (week 1),
and if so does the effect persist into the long-run (week 7)?

1. **Plot the Data.**  Roughly, there's a strong treatment effect one week
post-intervention that weakens by week 7. We'll try capturing this pattern in a
small world model. We focus on IL1-a, the cytokine with the fewest measurements
below the limit of detection.

   ```{r}
#| label: load-cytokines-all
#| fig-height: 4.5
#| fig-width: 8
ids <- read_csv("data/00_sample_ids_UKZN_workshop_2023.csv")
cytokines <- read_csv("data/03_elisa_cytokines_UKZN_workshop_2023.csv") |>
    left_join(ids)

ggplot(cytokines) +
    geom_line(aes(time_point, log(conc), group = pid, col = arm)) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_discrete(expand = c(0, 0)) +
    facet_wrap(~ reorder(cytokine, -conc), nrow = 2)
   ```

1. **Working Assumptions.** Luminex gives a continuous readout of each protein's
concentration, which suggests a linear regression model.

   The Gaussian error assumption suites Luminex well: many small, unrelated
   factors in the measurement or biology could make the fluorescence slightly
   brighter or dimmer. This also suggests that we model log-concentration
   instead of raw concentration, since errors are likely multiplicative rather
   than additive.

1. **Model Checks.** The code below fits a regression model of log-concentration
on treatment status. The treatment effect is significant, but we know to
interpret a model only after checking its fidelity.

   ```{r}
#| label: lm-cytokine
cytokines <- cytokines |>
    filter(cytokine == "IL-1a")

fit <- lm(log(conc) ~ arm, data = cytokines)
modelsummary(fit, shape = term ~ statistic, gof_map = NA)
   ```

   To this end, let's simulate from the fitted model (full in `helpers.R`).

   ```{r}
#| label: cytokine-wide
#| fig-height: 3.5
#| fig-width: 6.5
sim_data <- simulate_cytokine(fit)
plot_cytokine(sim_data)
   ```

   Reassuringly, the overall concentrations are in the correct range, and the
   control group has larger concentrations compared to treatment, matching the
   real data. But there is an issue -- in the real data, most blue curves
   decrease at week 1, but in our simulation about half the curves increase.

1. **Revised Models.** We shouldn't be surprised the model missed the week 1
decrease: we gave it no temporal information.  One potential fix is the formula
`~ arm + time_point`.

   But considering the large world data story, this model isn't quite right.
   We don't expect IL1-a concentration to change just because the clock reads
   week 1 or 7. We want the treatment effect itself to depend on the week. This
   is an interaction, `~ arm + arm : timepoint`.

1. The fitted model is printed below. The way to interpret the parameters is
that the expected log-concentrations are,

   - Placebo at baseline: 4.66
   - Treatment at baseline: 4.66 - 0.29
   - Placebo at week 1: 4.66 - 0.48
   - Treatment at week 1: 4.66 - 2.34
   ...

   In particular the large negative coefficient for the
   `armtreatment:time_pointweek_1` term captures our earlier observation that
   the treatment effect seems strongest at week 1.

   ```{r}
#| label: cytokines-interaction
fit <- lm(log(conc) ~ arm + arm : time_point, data = cytokines)
modelsummary(fit, shape = term ~ statistic, gof_map = NA)
   ```

   _Exercise: Does this model allow for a treatment effect at time zero? Based on what you know about the "large world" experiment, decide whether this is appropriate or if the model could be further improved._

1. The simulated data look better than before.

   ```{r}
#| label: cytokines-interaction-plot
#| fig-height: 3.5
#| fig-width: 6.5
sim_data <- simulate_cytokine(fit)
plot_cytokine(sim_data)
   ```

1. One more revision. The model doesn't account for longitudinality, which could
inflate our confidence.  It treats each sample as independent, when measurements
from the same participant are possibly dependent. The effective sample size lies
somewhere between the number of study participants (44) and measurements (132).

   We have the machinery to deal with this: a participant-level random effect,
   allowing a different baseline for each participant.

   ```{r}
fit <- lmer(log(conc) ~ arm + arm : time_point + (1 | pid), data = cytokines)
modelsummary(fit, shape = term ~ statistic, gof_map = NA)
   ```

1. Visually, the model doesn't seem much better. Indeed, the participant-level
variation (SD 0.15) is small compared to the within-sample variation (0.81).
Nonetheless, it's the safer model in practice, because by accounting for
within-participant dependence, we are being honest about the effective sample
size.

   ```{r}
#| label: cytokine-sim-re
#| fig-height: 3.5
#| fig-width: 6.5
sim_data <- simulate_cytokine(fit)
plot_cytokine(sim_data)
   ```

### Flow Cytometry Cell Compositions

1. **Biological Question.** Does the inflammation-reducing drug influence immune
cell composition at week 1 post-treatment?

1. Let's read in the data. Since our question focuses on `week_1`, we keep only
these sample IDs. The `right_join` then drops the other timepoints.

   ```{r}
#| label: load-flow-cytometry
ids <- read_csv("data/00_sample_ids_UKZN_workshop_2023.csv") |>
    filter(time_point == "week_1")
flow <- read_csv("data/04_flow_cytometry_UKZN_workshop_2023.csv") |>
    right_join(ids) |>
    mutate(arm = as.factor(arm))

head(flow)
   ```

1. **Plot the data.** We've reshaped the flow cytometry data into each sample's
cell type composition, shown as a stacked bar plot below.  Samples are sorted so
that those with lower cell type diversity appear on the left.

   ```{r}
#| label: plot-flow-cytometry
#| fig-height: 3.5
#| fig-width: 7.5
reshaped_flow <- flow |>
    pivot_longer(matches("cd|neutro"), names_to = "cell_type", values_to = "count") |>
    group_by(sample_id) |>
    mutate(proportion = count / sum(count))

ggplot(reshaped_flow) +
        geom_col(aes(reorder(sample_id, proportion, entropy), proportion, fill = cell_type), col = "black", width = 1, linewidth = 0.3) +
        facet_wrap(~ arm, scales = "free_x") +
        scale_y_continuous(expand = c(0, 0)) +
        labs(
            x = "Sample ID",
            fill = "Cell Type",
            y = "Proportion"
        ) +
        theme(axis.text.x = element_text(angle = 90))
   ```

   _Exercise: Replace `proportion` with `count` in the `geom_col` command above. What do you see? What could be some corresponding "large world" explanations?_

   The main question is whether there is a systematic difference in cell
   compositions between the treatment and control at week 1. If there is a
   difference, which cell types are responsible?

1. Before we build our small world model to answer this, it's worth considering
the "large world" data story,

   - There are on the order of a trillion immune cells (of various cell types)
   circulating through a typical body at any moment. Some cell types are much
   rarer than others, and there are more cell types than we would be able to
   assay with our flow cytometer. Cells are being created and destroyed at all
   times and they are not evenly spatially distributed, instead having
   concentrations shifting in response to signalling molecules and random drift.
   - At a doctor's office, a CVL is collected, catching some of the cells from
   the previous step that just happened to be present at the right place and
   time. A technician takes the collected sample and freezes it, killing all the
   cells present.
   - After a long time in the freezer, the sample is prepared for the flow
   cytometer. Most (but not all) cell-surface markers are bound by labeled
   antibodies, and the device records the associated profiles. All cells are
   destroyed in the process; from here on out it's just data.
   - A computational biologist gates the raw cytometer output into distinct cell
   types. Some don't fall anywhere we recognize and so are dropped from
   subsequent analysis. The data are saved to a CSV file and uploaded to the
   workshop website, living on a server in a data center in Greenland.  Three
   years later, you download a copy of that data to your computer.

   This is a big world, and our final file is an imperfect view of the
   participants' original immune landscapes. Sampling variation and measurement
   error enter at each step.

1. **Working Assumptions.** Our small world is quite simple.
Consider the population of immune cells in one participant. When we draw a
sample, we "catch" a mix of CD3- and other cell types.  With only two groups
(CD3- and "other") and only totals observed per participant, this is a natural
setting for a binomial model.

   ```{r}
#| label: fit-binom-model
flow <- flow |>
    mutate(
        total = rowSums(across(2:10)),
        proportion = cd3_negative / total
    )

fit <- glm(proportion ~ 1, family = binomial(), weights = total, data = flow)
modelsummary(fit, shape = term ~ statistic, gof_map = NA)
   ```

   After the inverse logistic transformation, the estimate is just the fraction
   of observed cells assigned to CD3-.

   ```{r}
#| label: transform-logit
plogis(-4.26)
sum(flow$cd3_negative) / sum(flow$total)
   ```

1. **Model Checks.** We simulate new data using the fitted model. The
proportions are the CD3- compositions of hypothetical samples.

   ```{r}
#| label: simulate-counts
y_sim <- simulate(fit)$sim_1
head(y_sim)
   ```

   In the simulated data, the CD3- proportions are close to but not exactly
   those in the real data, which reflects the randomness of the binomial model.

1. We've written a helper, `fit_cytof`, that loops this model over all cell
types and simulates counts for each.  Visualizing the simulated data, two issues
stand out,

   - The simulated proportions are very close to the estimated proportions
   across all the samples.  The binomial model has much lower variability than
   observed in the actual data.
   - Our model doesn't allow the fitted cell-type proportions to vary across
   treatment and control.

   We'll start with the second issue, because it's easier to address.  Of
   course, we'll need to answer the first before trusting this model.

   ```{r}
#| label: simulate-flow-counts
#| fig-height: 3.5
#| fig-width: 7.5
m <- fit_cytof(proportion ~ 1)
counts <- simulate_counts(m)
plot_counts(counts)
   ```

1. **Revised Models.** To let cell-type proportions depend on treatment, we
modify the formula from `proportion ~ 1` to `proportion ~ arm`.

   ```{r}
#| label: groupwise-flow-fit
m <- fit_cytof(proportion ~ arm)
   ```

1. The estimated coefficients with treatment status as a covariate are printed
below. The row for `cd3_negative` means that placebo samples have about
`plogis(-3.98) = 0.018` of their compostion from CD3- cells. Under treatment
this decreases to about `plogis(-3.98 - 0.578) = 0.010`.

   ```{r}
coef_df <- map_dfr(m, coef, .id = "cell_type")
coef_df
   ```

   We could in principle interpret changes in composition due to treatment from
   this model. But we shouldn't trust this model yet because it still fails to
   account for sample-to-sample variability.

   ```{r}
#| label: variable-cell-counts
#| fig-height: 3.5
#| fig-width: 7.5
counts <- simulate_counts(m)
plot_counts(counts)
   ```

1. To account for sample-to-sample differences, we add a random effect.
Since the sample labels are kept the `sample_id` column, we modify the formula to

   ```
    proportion ~ arm + (1 | sample_id)
   ```

   and switch from `glm` to `glmer`.

1. The new simulated data are much more realistic. They still don't quite have
the smooth variation in composition across samples seen in the real data
-- this could be introduced through latent factors or modeling
correlation across cell types. Even so, the current model is often sufficient
real data analyses, and underlies the `diffcyt` Bioconductor package @Weber2019.

   ```{r}
#| label: account-random-effects
#| fig-height: 3.5
#| fig-width: 7.5
fits <- fit_cytof(proportion ~ arm + (1 | sample_id), glmer)
counts <- simulate_counts(fits)
plot_counts(counts)
   ```

1. The model output shows how the treatment affects each cell type
(`armtreatment`) and how much varies across sample (`sample_id` `Std. Dev`). For
example, under treatment the CD3- composition tends to decrease relative to the
other cell types.

   ```{r}
#| label: composition-model-summary
modelsummary(fits[["cd3_negative"]], shape = term ~ statistic, gof_map = NA)
   ```

   _Exercise: According to the fitted model, what is the typical proportion of CD3- cells in the control and treatment groups? Hint: Use the `plogis` function applied to the right choice of values under `Estimate`._
