---
title: "gtsummary and ggpubr for hypothesis testing"
subtitle: "Workshop 1, Module 6"
format: qmdrslides-revealjs
editor: source
execute:
  echo: true
  message: false
  warning: false
---

<!-- Task 0.1 decision (2026-05-25): M3/M4 audit has NOT landed.
     Beat B4 in Phase 3 uses Variant B — prepend the 2-slide cytokine intro. -->

```{r setup, include=FALSE}
library(tidyverse)
library(countdown)
library(here)
library(gtsummary)
library(ggpubr)

table_01 <- read_csv(
  here("datasets/instructional_dataset/01_participant_metadata_UKZN_workshop_2023.csv"),
  col_types = cols(sex = col_character())
)
table_02 <- read_csv(here("datasets/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))
table_00 <- read_csv(here("datasets/instructional_dataset/00_sample_ids_UKZN_workshop_2023.csv"))
table_03 <- read_csv(here("datasets/instructional_dataset/03_elisa_cytokines_UKZN_workshop_2023.csv"))
```

## Goals for this module

By the end of this module you can:

- read an analysis plan and pick the gtsummary or ggpubr call for each row
- produce a Table 1 with `tbl_summary() |> add_p()` and override the default summary statistic and test per column
- produce a Table 2 (outcome by arm) and the matching boxplot with `stat_compare_means()`
- screen many outcomes at once with `facet_wrap()` and one `stat_compare_means()` per panel
- recognise when a paired test is the right test
- report an adjusted analysis with `lm() |> tbl_regression()`

## The analysis plan comes first

Every paper has an analysis plan. Each row is one question.

- Who is in the study? → `tbl_summary()`
- Did randomization balance the arms? → `tbl_summary() |> add_p()` (sanity check)
- Does the outcome differ between arms? → same machinery, or a boxplot + `stat_compare_means()`
- Did the outcome change pre vs post in the same person? → paired test
- Does the effect hold after adjusting for a covariate? → `lm() |> tbl_regression()`

The whole module walks this list. The tool falls out of the question.

## Who is in the study?

```{r}
table_01 |>
  select(-pid, -sex) |>
  tbl_summary(by = arm)
```

Defaults: continuous columns get *median (IQR)*, categorical columns get *counts (%)*. No p-values yet.

## Look at the columns before you trust the summary

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3
library(patchwork)

p1 <- ggplot(table_01, aes(age, fill = arm)) +
  geom_histogram(binwidth = 2, position = "identity", alpha = 0.6) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal(base_size = 11)

p2 <- ggplot(table_01, aes(smoker, fill = arm)) +
  geom_bar(position = position_dodge(0.8)) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal(base_size = 11)

p3 <- ggplot(table_01, aes(education, fill = arm)) +
  geom_bar(position = position_dodge(0.8)) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))

(p1 + p2 + p3) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

The Table 1 is a summary of these distributions. Always know what they look like before you trust the summary.

## Distribution drives two decisions

The decision to describe a column as *median (IQR)* vs *mean (SD)* is the same decision as choosing Wilcoxon vs t-test. Both follow from how skewed the column is.

- skewed continuous → median (IQR) + Wilcoxon
- symmetric continuous → mean (SD) + t-test
- sparse categorical (small cells) → counts (%) + Fisher exact
- dense categorical → counts (%) + chi-squared

gtsummary lets you set both with parallel `statistic = list(...)` and `test = list(...)` arguments.

## Override the defaults for one column

```{r}
table_01 |>
  select(-pid, -sex) |>
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})"),
    digits = list(age ~ 1)
  ) |>
  add_p(test = list(age ~ "t.test"))
```

`statistic = list(age ~ "{mean} ({sd})")` controls how age is shown. `test = list(age ~ "t.test")` controls how it is tested. The same `<column> ~ <choice>` shape works for both.

To do this for every continuous column at once: `all_continuous() ~ "{mean} ({sd})"`.

## A read_csv gotcha worth knowing

```{r}
#| eval: false
# Without col_types — sex parses as logical FALSE
table_01_broken <- read_csv(
  here("data/instructional_dataset/01_participant_metadata_UKZN_workshop_2023.csv")
)

table_01_broken |> count(sex)
#> # A tibble: 1 × 2
#>   sex       n
#>   <lgl> <int>
#> 1 FALSE    44
```

```{r}
#| eval: false
# With col_types — sex parses as character "F"
table_01 <- read_csv(
  here("data/instructional_dataset/01_participant_metadata_UKZN_workshop_2023.csv"),
  col_types = cols(sex = col_character())
)

table_01 |> count(sex)
#> # A tibble: 1 × 2
#>   sex       n
#>   <chr> <int>
#> 1 F        44
```

`read_csv()` sees single-character "F" values and infers boolean. Always look at column types after reading; specify them when in doubt.

## Polish for publication

```{r}
table_01 |>
  select(-pid, -sex) |>
  tbl_summary(by = arm) |>
  add_p() |>
  add_overall() |>
  bold_labels() |>
  modify_header(label ~ "**Characteristic**")
```

`add_overall()` adds a column for the full sample. `bold_labels()` makes row labels bold. `modify_header()` renames columns. A few lines turn a default table into a publication-ready one.

## Exercise 1 {.your-turn}

Open `notebooks/06-table-one.qmd` in your Codespace.

<br>

Group A: agentic. Group B: manual.

<br>

```{r}
countdown::countdown(30)
```
