---
title: "gtsummary and ggpubr for hypothesis testing"
subtitle: "Workshop 1, Module 6"
format: qmdrslides-revealjs
editor: source
execute:
  echo: true
  message: false
  warning: false
---

<!-- Task 0.1 decision (2026-05-25): M3/M4 audit has NOT landed.
     Beat B4 in Phase 3 uses Variant B — prepend the 2-slide cytokine intro. -->

```{r setup, include=FALSE}
library(tidyverse)
library(countdown)
library(here)
library(gtsummary)
library(ggpubr)

table_01 <- read_csv(
  here("datasets/instructional_dataset/01_participant_metadata_UKZN_workshop_2023.csv"),
  col_types = cols(sex = col_character())
)
table_02 <- read_csv(here("datasets/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))
table_00 <- read_csv(here("datasets/instructional_dataset/00_sample_ids_UKZN_workshop_2023.csv"))
table_03 <- read_csv(here("datasets/instructional_dataset/03_elisa_cytokines_UKZN_workshop_2023.csv"))
```

## Goals for this module

By the end of this module you can:

- read an analysis plan and pick the gtsummary or ggpubr call for each row
- produce a Table 1 with `tbl_summary() |> add_p()` and override the default summary statistic and test per column
- produce a Table 2 (outcome by arm) and the matching boxplot with `stat_compare_means()`
- screen many outcomes at once with `facet_wrap()` and one `stat_compare_means()` per panel
- recognise when a paired test is the right test
- report an adjusted analysis with `lm() |> tbl_regression()`

## The analysis plan comes first

Every paper has an analysis plan. Each row is one question.

- Who is in the study? → `tbl_summary()`
- Did randomization balance the arms? → `tbl_summary() |> add_p()` (sanity check)
- Does the outcome differ between arms? → same machinery, or a boxplot + `stat_compare_means()`
- Did the outcome change pre vs post in the same person? → paired test
- Does the effect hold after adjusting for a covariate? → `lm() |> tbl_regression()`

The whole module walks this list. The tool falls out of the question.

## Refine your plan with an LLM

Before you write any code, paste your analysis plan into a chat with an LLM and ask:

- which questions are ambiguous?
- which comparisons did I forget?
- which defaults am I accepting without thinking?
- where does my plan slip into causal language without earning it?

The LLM does not know your study design, your hypotheses, or your prior literature. It can only push back on what you wrote. The plan stays yours.

## Who is in the study?

```{r}
table_01 |>
  select(-pid, -sex) |>
  tbl_summary(by = arm)
```

Defaults: continuous columns get *median (IQR)*, categorical columns get *counts (%)*. No p-values yet.

## Mean (SD) and median (IQR), side by side

The Table 1 above used median (IQR) by default. Here is what each summary looks like on two of the columns we will work with.

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3.5
library(patchwork)

summary_overlay <- function(d, x_var, title) {
  v <- d[[x_var]]
  stats <- tibble(
    label = c("mean (SD)", "median (IQR)"),
    center = c(mean(v), median(v)),
    lo = c(mean(v) - sd(v), quantile(v, 0.25)),
    hi = c(mean(v) + sd(v), quantile(v, 0.75))
  )
  ggplot() +
    geom_jitter(data = d, aes(.data[[x_var]], y = "data"),
                height = 0.15, width = 0, alpha = 0.35, size = 1) +
    geom_pointrange(data = stats,
                    aes(x = center, xmin = lo, xmax = hi, y = label, color = label),
                    linewidth = 1.2, fatten = 5) +
    scale_y_discrete(limits = c("data", "mean (SD)", "median (IQR)")) +
    scale_color_brewer(palette = "Set1") +
    labs(title = title, x = x_var, y = NULL) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none")
}

p_sym <- summary_overlay(table_02, "ph", "ph — roughly symmetric")
p_skew <- summary_overlay(table_02, "crp_blood", "crp_blood — right-skewed")

p_sym + p_skew
```

For symmetric columns the two intervals agree. For skewed columns, the mean is pulled toward the tail; the IQR stays where most of the data sits.

## Look at the columns before you trust the summary

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3
library(patchwork)

p1 <- ggplot(table_01, aes(age, fill = arm)) +
  geom_histogram(binwidth = 2, position = "identity", alpha = 0.6) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal(base_size = 11)

p2 <- ggplot(table_01, aes(smoker, fill = arm)) +
  geom_bar(position = position_dodge(0.8)) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal(base_size = 11)

p3 <- ggplot(table_01, aes(education, fill = arm)) +
  geom_bar(position = position_dodge(0.8)) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))

(p1 + p2 + p3) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

The Table 1 is a summary of these distributions. Always know what they look like before you trust the summary.

## Distribution drives two decisions

The decision to describe a column as *median (IQR)* vs *mean (SD)* is the same decision as choosing Wilcoxon vs t-test. Both follow from how skewed the column is.

- skewed continuous → median (IQR) + Wilcoxon
- symmetric continuous → mean (SD) + t-test
- sparse categorical (small cells) → counts (%) + Fisher exact
- dense categorical → counts (%) + chi-squared

gtsummary lets you set both with parallel `statistic = list(...)` and `test = list(...)` arguments.

## Override the defaults for one column

```{r}
table_01 |>
  select(-pid, -sex) |>
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})"),
    digits = list(age ~ 1)
  ) |>
  add_p(test = list(age ~ "t.test"))
```

`statistic = list(age ~ "{mean} ({sd})")` controls how age is shown. `test = list(age ~ "t.test")` controls how it is tested. The same `<column> ~ <choice>` shape works for both.

To do this for every continuous column at once: `all_continuous() ~ "{mean} ({sd})"`.

## A read_csv gotcha worth knowing

```{r}
#| eval: false
# Without col_types — sex parses as logical FALSE
table_01_broken <- read_csv(
  here("data/instructional_dataset/01_participant_metadata_UKZN_workshop_2023.csv")
)

table_01_broken |> count(sex)
#> # A tibble: 1 × 2
#>   sex       n
#>   <lgl> <int>
#> 1 FALSE    44
```

```{r}
#| eval: false
# With col_types — sex parses as character "F"
table_01 <- read_csv(
  here("data/instructional_dataset/01_participant_metadata_UKZN_workshop_2023.csv"),
  col_types = cols(sex = col_character())
)

table_01 |> count(sex)
#> # A tibble: 1 × 2
#>   sex       n
#>   <chr> <int>
#> 1 F        44
```

`read_csv()` sees single-character "F" values and infers boolean. Always look at column types after reading; specify them when in doubt.

## Polish for publication

```{r}
table_01 |>
  select(-pid, -sex) |>
  tbl_summary(by = arm) |>
  add_p() |>
  add_overall() |>
  bold_labels() |>
  modify_header(label ~ "**Characteristic**")
```

`add_overall()` adds a column for the full sample. `bold_labels()` makes row labels bold. `modify_header()` renames columns. A few lines turn a default table into a publication-ready one.

## Randomization protects the arm comparison

Randomization is what justifies the unadjusted `add_p()` between placebo and treatment.

It does not protect any other comparison.

If you ran `tbl_summary(by = smoker)` instead of `by = arm`, you would be comparing two non-randomized groups. Smokers and non-smokers differ in many ways the table does not show — age, socioeconomic status, the environments they live in, the things they were exposed to before the study.

This is the limit of what Table 1 can sanity-check. Comparing non-randomized groups validly requires causal-inference machinery — DAGs, propensity scores, sensitivity analyses — which is beyond this workshop's scope.

## Exercise 1 {.your-turn}

Open `notebooks/06-table-one.qmd` in your Codespace.

<br>

Group A: agentic. Group B: manual.

<br>

```{r}
countdown::countdown(30)
```

# Outcomes & Table 2 questions {.center}

## From describe to compare

Half A asked who is in the study.

Half B asks whether the outcome differs between arms, within participants over time, or after adjusting for a covariate.

Same dataset, new family of questions.

## Table 2: outcome by arm

```{r}
table_02 |>
  filter(time_point == "week_7") |>
  select(-pid, -time_point) |>
  tbl_summary(by = arm) |>
  add_p()
```

Three outcome rows, two arms. `add_p()` picks Wilcoxon for all three (the columns are non-normal at this n). The p-values are now real inferential answers, not sanity checks.

## Same question, visual form

```{r}
table_02 |>
  filter(time_point == "week_7") |>
  ggboxplot(x = "arm", y = "ph", fill = "arm", palette = "Set1") +
  stat_compare_means(method = "wilcox.test", label.x = 1.3)
```

`ggboxplot()` is ggpubr's themed boxplot wrapper. `stat_compare_means()` is a ggplot layer that adds the p-value. The table-form and figure-form of this test should give the same p-value, and they do.

## A new table: cytokines

```{r}
table_03 |> head()
```

`table_03` is *long format*: one row per (sample, cytokine). Each `sample_id` has one row per cytokine measured on it.

The `limits` column flags whether the measured concentration fell within the assay's detection range.

## Linking cytokines to participant data

```{r}
table_03 |>
  left_join(table_00, by = "sample_id") |>
  head()
```

`table_00` is the junction: it maps `sample_id` to `pid` + `time_point` + `arm`. One left_join gets us from cytokine values to study arm.

## Same idea, multiple outcomes at once

```{r}
#| fig-width: 9
#| fig-height: 6
cytokines_keep <- c("IL-1a", "IL-6", "IL-8", "IL-1b", "MIG", "IP-10")

cytokine_panel_df <- table_03 |>
  filter(cytokine %in% cytokines_keep, limits == "within limits") |>
  left_join(table_00, by = "sample_id") |>
  filter(time_point == "week_7")

ggplot(cytokine_panel_df, aes(arm, conc, fill = arm)) +
  geom_boxplot() +
  facet_wrap(~ cytokine, scales = "free_y") +
  stat_compare_means(method = "wilcox.test", label.x = 1.3) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

Six cytokines, one figure, one Wilcoxon p-value per panel.

## Multiple panels means multiple tests

Six panels means six p-values. Each is a separate hypothesis test.

Correcting for multiple testing (Bonferroni, FDR) belongs to Module 5. Here, just know it is a thing — a single panel with a small p-value among six is not the same kind of evidence as a single panel with a small p-value standing alone.

## Paired pre-post: visual

```{r}
#| fig-width: 8
#| fig-height: 4
pre_post_df <- table_02 |>
  filter(time_point %in% c("baseline", "week_7")) |>
  mutate(time_point = fct_relevel(time_point, "baseline", "week_7"))

ggplot(pre_post_df, aes(time_point, ph)) +
  geom_line(aes(group = pid), alpha = 0.4) +
  geom_point(aes(color = arm), size = 2) +
  facet_wrap(~ arm) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 12)
```

Each line is one participant's trajectory from baseline to week 7. Same data as the boxplot a few slides back. Different question: instead of comparing two groups, we are looking at change within each person.

## Paired pre-post: the test changes too

```{r}
#| fig-width: 8
#| fig-height: 4.5
ggplot(pre_post_df, aes(time_point, ph)) +
  geom_line(aes(group = pid), alpha = 0.4) +
  geom_point(aes(color = arm), size = 2) +
  facet_wrap(~ arm) +
  stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    comparisons = list(c("baseline", "week_7"))
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 12)
```

`paired = TRUE` tells `stat_compare_means()` that each row at "baseline" pairs with the same participant's row at "week_7". The test is a paired Wilcoxon (signed-rank).

The data did not change. The unit of comparison did — each participant's *delta* is now the unit. So the test changed too.

## Adjusted analysis with tbl_regression

```{r}
outcome_table <- table_02 |>
  filter(time_point == "week_7") |>
  left_join(table_01, by = c("pid", "arm"))

lm(ph ~ arm + age + smoker, data = outcome_table) |>
  tbl_regression()
```

`tbl_regression()` turns a model fit into a publication-ready coefficient table. Each row is one coefficient with its estimate, confidence interval, and p-value.

In a randomized trial, adjustment is mostly cosmetic: the unadjusted t-test is the primary analysis. In a case-control study or any non-randomized design, you genuinely need this.

## The analysis plan, recap

- one outcome, two groups, cross-sectional → `tbl_summary + add_p` or `ggboxplot + stat_compare_means`
- many outcomes, two groups → `facet_wrap` + `stat_compare_means`
- same person pre vs post → paired test on a line plot
- need to adjust for a covariate → `lm` + `tbl_regression`

The right tool depends on the question you are asking.

## Exercise 2 {.your-turn}

Open `notebooks/06-figures-with-pvalues.qmd` in your Codespace.

<br>

The modes flip: whoever was agentic in Exercise 1 is manual now, and vice versa.

<br>

```{r}
countdown::countdown(30)
```
