---
title: "The vaginal immune environment across the menstrual cycle"
subtitle: "An example group-project presentation · Group B"
format: qmdrslides-revealjs
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(here)
library(gtsummary)
library(ggpubr)
library(lme4)
library(lmerTest)
library(broom.mixed)

theme_set(theme_minimal(base_size = 18))
pal <- "Set1"
cycle <- c("week_prior", "onset", "end_bleeding", "week_post")

dd <- here("datasets", "group_b_menstruation")
ids  <- read_csv(file.path(dd, "00_sample_ids_period.csv"))
meta <- read_csv(file.path(dd, "01_participant_metadata_period.csv"))
lum  <- read_csv(file.path(dd, "02_luminex_period.csv"))
flow <- read_csv(file.path(dd, "03_flow_cytometry_period.csv"))

ids <- ids |>
  mutate(
    time_point = factor(time_point, levels = cycle),
    arm        = fct_relevel(arm, "no_birth_control", "birth_control")
  )

cyt  <- lum  |> left_join(ids, by = "sample_id")
flow <- flow |> left_join(ids, by = "sample_id") |>
  mutate(prop_neut = neutrophils / cd45_positive)
```

# Part 1 · The dataset {background-color="#3b1b4b"}

## The study

An **observational** study of 27 women, 15 using hormonal birth control and 12
not. Each woman was sampled at **four points across one menstrual cycle**:

::: {style="font-size: 0.9em"}
the week before bleeding → onset → end of bleeding → the week after
:::

From each sample we have:

- **Luminex** — 8 cytokines (inflammation markers)
- **flow cytometry** — counts of immune cell types (neutrophils, T cells, …)
- participant metadata — age, PCOS status, period product

Two questions: **how does the vaginal immune environment shift across the
cycle, and does birth control change it?**

## Who is in the study?

```{r}
#| label: table-one
meta |>
  select(arm, age, pcos_status, period_product) |>
  tbl_summary(
    by = arm,
    label = list(
      age ~ "Age (years)",
      pcos_status ~ "PCOS status",
      period_product ~ "Period product"
    )
  ) |>
  add_p() |>
  add_overall() |>
  bold_labels() |>
  modify_header(label ~ "**Characteristic**")
```

::: {style="font-size: 0.7em"}
This is a small cohort, so categorical rows use Fisher's exact test.
:::

## How the study is laid out

```{r}
#| label: design-plot
#| fig-width: 12
#| fig-height: 6
ids |>
  ggplot(aes(time_point, fct_rev(pid), colour = arm)) +
  geom_line(aes(group = pid), colour = "grey75") +
  geom_point(size = 3) +
  facet_wrap(~ arm, scales = "free_y") +
  scale_colour_brewer(palette = pal) +
  labs(x = NULL, y = "Participant",
       title = "Four samples per woman, spanning one cycle") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1),
        axis.text.y = element_blank())
```

## Which cytokines can we use?

```{r}
#| label: cytokine-detection
#| fig-width: 11
#| fig-height: 5
cyt |>
  group_by(cytokine) |>
  summarise(detected = mean(limits == "not_censored"), .groups = "drop") |>
  ggplot(aes(detected, fct_reorder(cytokine, detected))) +
  geom_col(fill = "#984ea3") +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "grey40") +
  scale_x_continuous(labels = scales::label_percent(), limits = c(0, 1)) +
  labs(x = "Measurements above the detection limit", y = NULL,
       title = "IL-1a is detected almost everywhere (dashed line = 75%)")
```

## What the immune cells look like

```{r}
#| label: flow-overview
#| fig-width: 12
#| fig-height: 5.5
flow |>
  transmute(time_point, sample_id,
    Neutrophils  = neutrophils,
    `T cells (CD3+)` = cd3_positive,
    `CD3- cells`     = cd3_negative,
    Other            = pmax(cd45_positive - neutrophils - cd3_positive - cd3_negative, 0)
  ) |>
  pivot_longer(-c(time_point, sample_id), names_to = "cell_type", values_to = "count") |>
  group_by(time_point, cell_type) |>
  summarise(count = sum(count), .groups = "drop_last") |>
  mutate(prop = count / sum(count)) |>
  ggplot(aes(time_point, prop, fill = cell_type)) +
  geom_col(width = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::label_percent(), expand = c(0, 0)) +
  labs(x = NULL, y = "Share of CD45+ immune cells", fill = NULL,
       title = "Neutrophils dominate, and their share shifts during bleeding") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

# Part 2 · Three short analyses {background-color="#3b1b4b"}

## Three questions

1. Does **inflammation** (IL-1a) rise during menstruation?
2. Does the **neutrophil share** of immune cells change across the cycle?
3. Does **birth control** shift the cytokine milieu?

# Story 1 · Inflammation across the cycle {background-color="#3a2a4d"}

## The question

Bleeding is a tissue-remodelling event, so we expect inflammation to track it.

**Does IL-1a rise during menstruation and settle afterwards?** Each woman is
measured four times, so we fit a model with a **per-participant random effect**
and let the week-before-bleeding be the reference point.

```{r}
#| label: story1-motivation
#| fig-width: 10
#| fig-height: 3.4
cyt |>
  filter(cytokine == "IL-1a") |>
  ggplot(aes(time_point, conc, group = pid)) +
  geom_line(alpha = 0.3, colour = "#e41a1c") +
  scale_y_log10() +
  labs(x = NULL, y = "IL-1a (log)") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

## The code

```r
il1 <- cyt |> filter(cytokine == "IL-1a")

fit <- lmer(log(conc) ~ time_point + (1 | pid), data = il1)

tbl_regression(fit, intercept = TRUE)
```

`time_point` is a factor ordered from the week before bleeding, so every
coefficient reads as a change *relative to that baseline week*. `(1 | pid)`
accounts for the four correlated samples per woman.

## The table

```{r}
#| label: story1-table
il1 <- cyt |> filter(cytokine == "IL-1a")

fit <- lmer(log(conc) ~ time_point + (1 | pid), data = il1)

tbl_regression(fit, intercept = TRUE) |>
  bold_labels() |>
  modify_header(label ~ "**Term**")
```

## What it means

On the log scale, relative to the week before bleeding:

- **Onset: +2.3** — roughly a tenfold jump in IL-1a (p < 0.001).
- **End of bleeding: +2.0** — still strongly elevated.
- **Week after: +0.3** — back to baseline (p = 0.4).

Inflammation rises sharply with bleeding and resolves within a week. The
cytokine milieu is a moving target that depends entirely on *when* in the cycle
you sample — a key lesson for designing studies like this one.

# Story 2 · Neutrophils across the cycle {background-color="#3a2a4d"}

## The question

The Part 1 overview hinted that neutrophils' share of immune cells changes
during bleeding. Let's test it properly.

The outcome is a **proportion of counted cells**, so the natural model is
binomial — exactly the flow-cytometry recipe from the modeling lecture. Because
the cell counts are huge, a plain binomial is wildly overconfident, so we add a
**per-sample random effect** to capture the real sample-to-sample spread.

## The code

```r
flow <- flow |> mutate(other = cd45_positive - neutrophils)

fit <- glmer(
  cbind(neutrophils, other) ~ time_point + (1 | pid) + (1 | sample_id),
  family = binomial(), data = flow
)

# the figure: observed neutrophil share across the cycle
ggboxplot(flow, x = "time_point", y = "prop_neut",
          fill = "time_point", palette = "Set1") +
  stat_compare_means(method = "wilcox.test", ref.group = "week_prior",
                     label = "p.signif")
```

## The figure

```{r}
#| label: story2-figure
#| fig-width: 11
#| fig-height: 5.6
ggboxplot(flow, x = "time_point", y = "prop_neut",
          fill = "time_point", palette = pal, add = "jitter") +
  stat_compare_means(method = "wilcox.test", ref.group = "week_prior",
                     label = "p.signif",
                     label.y = 1.02, size = 6) +
  scale_y_continuous(labels = scales::label_percent(),
                     expand = expansion(mult = c(0.05, 0.18))) +
  labs(x = NULL, y = "Neutrophils as % of CD45+ cells") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

::: {style="font-size: 0.7em"}
Stars compare each timepoint to the week before bleeding (ns = not significant).
:::

## What it means

```{r}
#| label: story2-model
flow <- flow |> mutate(other = cd45_positive - neutrophils)
fit <- glmer(cbind(neutrophils, other) ~ time_point + (1 | pid) + (1 | sample_id),
             family = binomial(), data = flow,
             control = glmerControl(optimizer = "bobyqa"))
fit |>
  tbl_regression(exponentiate = FALSE, intercept = FALSE) |>
  bold_labels() |>
  modify_header(label ~ "**Term**")
```

- The neutrophil share **dips at onset and end of bleeding** and is highest
  again the week after (model p ≈ 0.004 and 0.01).
- The per-sample random effect was essential: without it the model treats
  hundreds of thousands of cells as independent coin flips and reports
  impossibly tiny p-values.
- Participant-to-participant variation was near zero — the cycle stage, not the
  individual, drives the neutrophil share.

# Story 3 · Does birth control matter? {background-color="#3a2a4d"}

## The question

Hormonal contraception changes the cervicovaginal environment, so we might
expect it to leave a cytokine signature.

**Does any cytokine differ between the birth-control and no-birth-control
arms?** We screen the well-detected cytokines and correct for multiple testing.

## The code

```r
keep <- c("IL-1a", "IP-10", "MIP-3a", "MIG")   # detected in >75% of samples

panel <- cyt |> filter(limits == "not_censored", cytokine %in% keep)

ggboxplot(panel, x = "arm", y = "conc", fill = "arm",
          palette = "Set1", facet.by = "cytokine") +
  scale_y_log10(expand = expansion(mult = c(0.05, 0.2))) +
  stat_compare_means(method = "wilcox.test", label = "p.format")
```

## The figure

```{r}
#| label: story3-figure
#| fig-width: 12
#| fig-height: 6
keep <- c("IL-1a", "IP-10", "MIP-3a", "MIG")
panel <- cyt |> filter(limits == "not_censored", cytokine %in% keep)

ggboxplot(panel, x = "arm", y = "conc", fill = "arm",
          palette = pal, facet.by = "cytokine", outlier.shape = NA) +
  scale_y_log10(expand = expansion(mult = c(0.05, 0.2))) +
  stat_compare_means(method = "wilcox.test", label = "p.format",
                     label.x = 1.2, size = 4.5) +
  labs(x = NULL, y = "Concentration (log scale)") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 15, hjust = 1))
```

## What it means

```{r}
#| label: story3-table
cyt |>
  filter(limits == "not_censored", cytokine %in% keep) |>
  group_by(cytokine) |>
  summarise(n = n(),
            p = wilcox.test(conc ~ arm)$p.value,
            .groups = "drop") |>
  mutate(p_BH = p.adjust(p, "BH")) |>
  arrange(p) |>
  mutate(across(c(p, p_BH), ~ scales::label_pvalue()(.x))) |>
  knitr::kable(col.names = c("Cytokine", "n", "p", "p (BH)"))
```

No cytokine survives correction — not even before it. In this cohort, **the
menstrual cycle is a far stronger driver of the immune environment than birth
control is.** A null result, honestly reported, is still a finding.

# Wrap-up {background-color="#3b1b4b"}

## What we found

- **Inflammation rises and falls with bleeding** — IL-1a jumps ~10x at onset
  and resets within a week.
- **Neutrophils' share of immune cells dips during bleeding**, then rebounds.
- **Birth control left no detectable cytokine signature** against that much
  larger cyclic signal.

## The takeaway

Timing dominates. Any study of the vaginal immune environment has to account
for cycle stage before it can ask about anything else.

Every table and figure here came from the workshop's own toolkit —
`tidyverse`, `gtsummary`, `ggpubr`, and `lme4` — built with an LLM assistant,
with the scientific judgement kept firmly in human hands.
