---
title: "Yogurt & the vaginal microbiome"
subtitle: "An example group-project presentation · Group A"
format: qmdrslides-revealjs
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(here)
library(gtsummary)
library(ggpubr)
library(lme4)
library(lmerTest)
library(broom.mixed)

theme_set(theme_minimal(base_size = 18))
pal <- "Set1"

dd <- here("datasets", "group_a_yogurt")
ids  <- read_csv(file.path(dd, "00_sample_ids_yogurt.csv"))
meta <- read_csv(file.path(dd, "01_participant_metadata_yogurt.csv"))
qpcr <- read_csv(file.path(dd, "02_qpcr_results_yogurt.csv"))
lum  <- read_csv(file.path(dd, "03_luminex_results_yogurt.csv"))

ids <- ids |>
  mutate(
    time_point = fct_relevel(time_point, "baseline", "after_antibiotic"),
    arm        = fct_relevel(arm, "unchanged_diet", "yogurt")
  )

# qPCR joined to study design, long over the three targets
qpcr_long <- qpcr |>
  left_join(ids, by = "sample_id") |>
  pivot_longer(starts_with("qpcr_"), names_to = "target", values_to = "copies") |>
  mutate(target = recode(target,
    qpcr_bacteria  = "Total bacteria",
    qpcr_crispatus = "L. crispatus",
    qpcr_iners     = "L. iners"
  ))

# Cytokines joined to study design
cyt <- lum |>
  left_join(ids, by = "sample_id")
```

# Part 1 · The dataset {background-color="#1b3a4b"}

## The study

A **randomized controlled trial** in 51 women. Everyone received a course of
antibiotics; half were randomized to eat yogurt daily, half kept their usual
diet.

Each woman was sampled **twice** — once at baseline, once after the antibiotic
course — and three kinds of measurement were taken from each sample:

- **qPCR** — absolute load of total bacteria, *L. crispatus*, and *L. iners*
- **Luminex** — concentrations of 10 cytokines (inflammation markers)
- participant metadata — age, contraception, education

The question: **does yogurt shape how the vaginal microbiome recovers from
antibiotics?**

## Who is in the study? {auto-animate="true"}

```{r}
#| label: table-one
meta |>
  select(arm, age, birth_control, education, days_since_last_sex) |>
  tbl_summary(
    by = arm,
    label = list(
      age ~ "Age (years)",
      birth_control ~ "Contraception",
      education ~ "Education",
      days_since_last_sex ~ "Days since last sex"
    )
  ) |>
  add_p() |>
  add_overall() |>
  bold_labels() |>
  modify_header(label ~ "**Characteristic**")
```

## A note on balance

The arms are well matched on age and education. Contraception is the one
variable to keep in mind: Depo-Provera is roughly twice as common in the
yogurt arm. Hormonal contraception can influence the vaginal environment, so
it is a plausible confounder we would want to revisit.

## How the study is laid out

```{r}
#| label: design-plot
#| fig-width: 12
#| fig-height: 6
ids |>
  ggplot(aes(time_point, fct_rev(pid), colour = arm)) +
  geom_line(aes(group = pid), colour = "grey75") +
  geom_point(size = 3) +
  facet_wrap(~ arm, scales = "free_y") +
  scale_colour_brewer(palette = pal) +
  labs(x = NULL, y = "Participant", colour = "Arm",
       title = "Two samples per woman: before and after antibiotics") +
  theme(legend.position = "none", axis.text.y = element_blank())
```

## The antibiotic course reshapes the microbiome

```{r}
#| label: qpcr-overview
#| fig-width: 12
#| fig-height: 5.5
qpcr_long |>
  ggplot(aes(time_point, copies + 1, fill = time_point)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_line(aes(group = pid), alpha = 0.15) +
  facet_wrap(~ target) +
  scale_y_log10(labels = scales::label_log()) +
  scale_fill_brewer(palette = pal) +
  labs(x = NULL, y = "Gene copies / mL (log scale)",
       title = "Total bacteria fall; lactobacilli bloom") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

::: {style="font-size: 0.7em"}
qPCR spans many orders of magnitude, so we always read it on a log scale.
:::

## Which cytokines can we actually use?

```{r}
#| label: cytokine-detection
#| fig-width: 11
#| fig-height: 5
cyt |>
  group_by(cytokine) |>
  summarise(within = mean(limits == "within limits"), .groups = "drop") |>
  ggplot(aes(within, fct_reorder(cytokine, within))) +
  geom_col(fill = "#377eb8") +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "grey40") +
  scale_x_continuous(labels = scales::label_percent(), limits = c(0, 1)) +
  labs(x = "Measurements within the assay's detection range", y = NULL,
       title = "Five cytokines are reliably detected (dashed line = 75%)")
```

# Part 2 · Three short analyses {background-color="#1b3a4b"}

## Three questions

1. After antibiotics, do **lactobacilli rebound**, and does yogurt change the
   rebound?
2. Does the antibiotic course change **IL-1a**, and does the effect depend on
   the diet arm?
3. Across **all** cytokines, which ones shift from before to after?

# Story 1 · Lactobacillus rebound {background-color="#2a4d3a"}

## The question

After a course of antibiotics, total bacterial load drops sharply. The
clinically interesting question is what grows back.

*L. iners* and *L. crispatus* are the "good" lactobacilli associated with a
healthy vaginal environment. **Do they rebound after antibiotics, and does
eating yogurt change how much they rebound?**

We compare each woman to herself (baseline vs after), separately within each
arm, with a paired Wilcoxon test.

## The code

```r
qpcr_long |>
  ggplot(aes(time_point, copies + 1)) +
  geom_boxplot(aes(fill = time_point), outlier.shape = NA, alpha = 0.8) +
  geom_line(aes(group = pid), alpha = 0.2) +
  facet_grid(arm ~ target) +
  scale_y_log10(expand = expansion(mult = c(0.05, 0.15))) +
  scale_fill_brewer(palette = "Set1") +
  stat_compare_means(method = "wilcox.test", paired = TRUE,
                     label = "p.format", size = 4) +
  labs(x = NULL, y = "Gene copies / mL (log scale)")
```

A paired test pairs each woman's *after* sample with her own *baseline*
sample, so we measure within-person change, not differences between people.

## The figure

```{r}
#| label: story1-figure
#| fig-width: 12.5
#| fig-height: 6
qpcr_long |>
  ggplot(aes(time_point, copies + 1)) +
  geom_boxplot(aes(fill = time_point), outlier.shape = NA, alpha = 0.8) +
  geom_line(aes(group = pid), alpha = 0.2) +
  facet_grid(arm ~ target) +
  scale_y_log10(expand = expansion(mult = c(0.05, 0.15)),
                labels = scales::label_log()) +
  scale_fill_brewer(palette = pal) +
  stat_compare_means(method = "wilcox.test", paired = TRUE,
                     label = "p.format", size = 4) +
  labs(x = NULL, y = "Gene copies / mL (log scale)") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

## What it means

- **Total bacteria fall** after antibiotics, as expected.
- ***L. iners* blooms** — its median jumps roughly four orders of magnitude in
  both arms (p < 0.001). This is the dominant recovery signal.
- ***L. crispatus*** is undetectable in most women at both timepoints, so its
  test is underpowered.
- The rebound looks **the same with or without yogurt**: the *after* boxes sit
  at the same height in both rows.

The headline: antibiotics clear the field and *L. iners* repopulates it —
yogurt does not visibly steer that process.

# Story 2 · Modeling IL-1a {background-color="#2a4d3a"}

## The question

*L. iners* came back. Did inflammation come down with it?

**IL-1a** is the best-detected cytokine here, so we model it directly. We want
two things at once: the change after antibiotics, and whether that change
differs by arm. That calls for an **interaction model**, with a per-participant
random effect because each woman appears twice.

```{r}
#| label: story2-motivation
#| fig-width: 10
#| fig-height: 3.4
cyt |>
  filter(cytokine == "IL-1a") |>
  ggplot(aes(time_point, conc, group = pid, colour = arm)) +
  geom_line(alpha = 0.5) +
  facet_wrap(~ arm) +
  scale_y_log10() +
  scale_colour_brewer(palette = pal) +
  labs(x = NULL, y = "IL-1a (log)") +
  theme(legend.position = "none")
```

## The code

```r
il1 <- cyt |> filter(cytokine == "IL-1a")

fit <- lmer(
  log(conc) ~ arm + arm:time_point + (1 | pid),
  data = il1
)

tbl_regression(fit, intercept = TRUE)
```

`arm:time_point` lets the after-antibiotic change be different in each arm.
`(1 | pid)` says the two samples from one woman are correlated. We model
`log(conc)` because cytokine measurement error is multiplicative.

## The table

```{r}
#| label: story2-table
il1 <- cyt |> filter(cytokine == "IL-1a")

fit <- lmer(
  log(conc) ~ arm + arm:time_point + (1 | pid),
  data = il1
)

tbl_regression(fit, intercept = TRUE) |>
  bold_labels() |>
  modify_header(label ~ "**Term**")
```

## What it means

Reading the coefficients on the log scale:

- **At baseline the arms are level** (`arm yogurt` ≈ 0, p = 0.9) — randomization
  worked.
- **IL-1a falls after antibiotics in both arms**: about −0.63 in the
  unchanged-diet arm and −0.77 in the yogurt arm, both clearly significant.
- The two drops are **close in size**, so yogurt does not meaningfully change
  the inflammatory response.

Same conclusion as Story 1, reached a different way: the microbiome recovers
and inflammation subsides, with or without yogurt.

# Story 3 · The whole cytokine panel {background-color="#2a4d3a"}

## The question

IL-1a fell. Was it special, or does the whole inflammatory panel come down
after antibiotics?

We run the **same paired test on every well-detected cytokine at once**, then
correct for testing several hypotheses.

## The code

```r
panel <- cyt |> filter(limits == "within limits")

# one paired Wilcoxon per cytokine, then Benjamini-Hochberg
panel |>
  pivot_wider(id_cols = c(pid, cytokine),
              names_from = time_point, values_from = conc,
              values_fn = mean) |>
  drop_na() |>
  group_by(cytokine) |>
  summarise(p = wilcox.test(baseline, after_antibiotic,
                            paired = TRUE)$p.value, n = n()) |>
  mutate(p_BH = p.adjust(p, "BH")) |>
  arrange(p_BH)
```

## The figure

```{r}
#| label: story3-figure
#| fig-width: 12.5
#| fig-height: 6
# the five cytokines detected in >75% of samples (Part 1)
keep <- c("IL-1a", "IP-10", "IL-8", "MIG", "MIP-3a")

cyt |>
  filter(limits == "within limits", cytokine %in% keep) |>
  ggboxplot(x = "time_point", y = "conc", fill = "time_point",
            palette = pal, facet.by = "cytokine", outlier.shape = NA) +
  geom_line(aes(group = pid), alpha = 0.12) +
  scale_y_log10(expand = expansion(mult = c(0.05, 0.2))) +
  stat_compare_means(method = "wilcox.test", paired = TRUE,
                     label = "p.format", size = 3.6) +
  labs(x = NULL, y = "Concentration (log scale)") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

## What it means

```{r}
#| label: story3-table
cyt |>
  filter(limits == "within limits", cytokine %in% keep) |>
  pivot_wider(id_cols = c(pid, cytokine),
              names_from = time_point, values_from = conc,
              values_fn = mean) |>
  drop_na() |>
  group_by(cytokine) |>
  summarise(n = n(),
            p = wilcox.test(baseline, after_antibiotic, paired = TRUE)$p.value,
            .groups = "drop") |>
  mutate(p_BH = p.adjust(p, "BH")) |>
  arrange(p_BH) |>
  mutate(across(c(p, p_BH), ~ scales::label_pvalue()(.x))) |>
  knitr::kable(col.names = c("Cytokine", "n pairs", "p", "p (BH)"))
```

Five cytokines move together and all survive Benjamini-Hochberg correction.
The inflammatory panel comes down as a group after antibiotics — IL-1a was
representative, not exceptional.

# Wrap-up {background-color="#1b3a4b"}

## What we found

- Antibiotics drop total bacterial load; ***L. iners* repopulates** strongly.
- **Inflammation subsides** alongside that recovery, across the cytokine panel.
- **Yogurt left no visible fingerprint** on either process in this trial.

## A caveat and a next step

- Depo-Provera was imbalanced between arms; a fair yogurt comparison should
  adjust for it.
- Everything here used only the workshop's own toolkit — `tidyverse`,
  `gtsummary`, `ggpubr`, and `lme4` — assembled with an LLM assistant. The
  judgement about *which* test, *which* scale, and *what it means* stayed with
  us.
