---
title: "More data wrangling and visualization"
subtitle: "Workshop 1, Module 4"
format: qmdrslides-revealjs
editor: source
execute:
  echo: true
  message: false
  warning: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(countdown)
library(here)
library(ggbeeswarm)

table_01 <- read_csv(here("datasets/instructional_dataset/01_participant_metadata_UKZN_workshop_2023.csv"))
table_02 <- read_csv(here("datasets/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))
```

## Goals for this module

By the end of this module you can:

- name what a row represents *before and after* every verb in a pipeline
- collapse a table to one row per group with `group_by` + `summarise`
- attach data from another table with `left_join`
- filter by membership in another table with `semi_join`
- reshape between long and wide with `pivot_longer` and `pivot_wider`
- pick the right plot for a distribution (histogram, boxplot, quasirandom, violin)
- layer raw observations under a computed summary in one figure

## Module 3 preserved row meaning. Module 4 changes it.

Module 3's verbs all kept "what a row represents" the same: `filter` drops rows but each remaining row still means the same thing; `mutate` adds columns; `arrange` reorders.

Module 4's verbs change it. Or sometimes don't. You have to know which.

The whole module asks one question on every slide: **what does a row represent now?**

## The through-line: `table_02`

```{r}
table_02
```

132 rows. *Row = participant-visit* (44 participants × 3 visits).

We will transform this same table through the next 18 slides. Watch what each verb does to the row count, and to what a row represents.

## `group_by` is a state change, not a shape change

```{r}
table_02 |> group_by(pid)
```

Same 132 rows. *Row meaning unchanged.* What changed is that subsequent verbs now act *per group*.

## `summarise` collapses each group to one row

```{r}
table_02 |>
  group_by(pid) |>
  summarise(
    mean_ph = mean(ph),
    mean_nugent = mean(nugent_score)
  )
```

132 → 44 rows. *Row = participant.* Visits collapsed.

## Different grouping, different row meaning

```{r}
table_02 |>
  group_by(arm, time_point) |>
  summarise(mean_ph = mean(ph))
```

132 → 6 rows. *Row = arm-timepoint.* Same starting data, different grouping, different row meaning.

## You picked what a row means

The verb didn't decide.

When you wrote `group_by(pid)`, the row collapsed to a participant.

When you wrote `group_by(arm, time_point)`, it collapsed to an arm-timepoint cell.

The grouping is the design choice that decides what a row will mean after `summarise`.

## Quick check {.your-turn}

What does each row represent, and how many rows, after:

```r
table_01 |>
  group_by(smoker) |>
  summarise(n = n(), mean_age = mean(age))
```

Predict before the reveal.

```{r}
countdown::countdown(1, top = 0)
```

## `group_by` + `mutate` — same rows, derived by group

```{r}
table_02 |>
  group_by(arm) |>
  mutate(rank_ph_within_arm = rank(ph)) |>
  select(pid, arm, ph, rank_ph_within_arm) |>
  head()
```

Still 132 rows, *row = participant-visit.* The new column is computed within each arm. `group_by` doesn't always collapse; the verb after it decides.

## `left_join` — the mutating join

`left_join` adds columns from a second table by matching on a key. *Same row meaning, same row count* (when the join is well-formed).

![](images/join.svg)

## `left_join` — demo

```{r}
per_person <- table_02 |>
  group_by(pid) |>
  summarise(mean_ph = mean(ph), mean_nugent = mean(nugent_score))

per_person |> left_join(table_01, by = "pid")
```

44 rows in, 44 rows out. *Row = participant* (unchanged). New columns: `arm`, `smoker`, `age`, `education`, `sex`.

## `semi_join` — the filtering join

`semi_join` keeps rows in the left table that have a match in the right table. *No new columns.* Same row meaning, fewer rows.

![](images/joins-venn.svg)

## `semi_join` — demo

```{r}
smokers <- table_01 |> filter(smoker == "smoker")

per_person |> semi_join(smokers, by = "pid")
```

*Row = participant* (unchanged). Rows that aren't smokers are dropped.

## Watch out — many-to-many joins multiply rows

If both sides have duplicate keys, the join multiplies rows:

```{r}
left <- tibble(id = c(1, 1, 2), x = c("a", "b", "c"))
right <- tibble(id = c(1, 1, 2), y = c("p", "q", "r"))
left |> left_join(right, by = "id")
```

3 + 3 = ... 5 rows. A well-formed join from a tidy left table should *not* change your row count. If it does, your key isn't unique somewhere.

## Quick check {.your-turn}

What does each row represent, and how many rows, after:

```r
table_02 |>
  left_join(table_01, by = c("pid", "arm"))
```

(Hint: `arm` appears in both tables. Does that change anything?)

```{r}
countdown::countdown(1, top = 0)
```

## `pivot_longer` — from wide to long

Each measurement column becomes a row.

![](images/pivot.svg)

## `pivot_longer` — demo

```{r}
table_02 |>
  pivot_longer(
    cols = c(ph, nugent_score, crp_blood),
    names_to = "measurement",
    values_to = "value"
  )
```

132 → 396 rows. *Row = participant-visit-measurement.* Useful for faceted plots — we'll use this in Exercise 2.

## `pivot_wider` — from long to wide

The inverse: one row per entity with the previously-long values spread out as columns.

![](images/pivot.svg)

## `pivot_wider` — demo

```{r}
arm_timepoint <- table_02 |>
  group_by(arm, time_point) |>
  summarise(mean_ph = mean(ph))

arm_timepoint |>
  pivot_wider(
    names_from = time_point,
    values_from = mean_ph
  )
```

6 → 2 rows. *Row = arm.* Three new columns, one per visit. Useful for table-style display.

## Exercise 1 {.your-turn}

Open `notebooks/04-wrangling.qmd` in your Codespace.

<br>

Group A: agentic. Group B: manual.

<br>

```{r}
countdown::countdown(30)
```

# Visualization {.center}

## Marks that summarise, or marks arranged by density

Module 3 said *every row of the data becomes a mark.*

Module 4 extends it. Some plots count rows into bins. Some summarise many rows into one mark. Some keep every row visible but use the layout to reveal density.

Several ways to show a distribution.

## `geom_histogram` — bin and count

```{r}
ggplot(table_02, aes(ph)) +
  geom_histogram(binwidth = 0.2, fill = "navyblue")
```

One bar = count of rows in a bin. *Many rows per mark.* Bin width matters — try `binwidth = 0.05` and `binwidth = 1` to see how the picture changes.

## `geom_bar` for counting

```{r}
ggplot(table_01, aes(education)) +
  geom_bar(fill = "navyblue")
```

Same idea as `count(education) |> geom_col()`. `geom_bar` does the count internally; `geom_col` is for pre-counted data.

## `geom_boxplot` revisited

```{r}
ggplot(table_02, aes(arm, ph)) +
  geom_boxplot()
```

A five-number summary of many rows. Default choice for comparing distributions across many groups.

## `geom_quasirandom` — every row visible

```{r}
ggplot(table_02, aes(arm, ph)) +
  geom_quasirandom()
```

From `ggbeeswarm`. Each row is still one mark (1:1), but horizontal jitter is set by local density at that y-value. The cluster shape reveals the distribution without abstracting rows away.

## Density, briefly

The fat parts of a quasirandom plot are where density is high.

That is what *density* means. We will see an explicit smoothed version in a moment.

## Plotting your own summary

```{r}
ph_summary <- table_02 |>
  group_by(arm) |>
  summarise(mean_ph = mean(ph), sd_ph = sd(ph))

ggplot(ph_summary, aes(arm, mean_ph)) +
  geom_col(fill = "navyblue")
```

After you summarise, each row is one summary → one bar. Row-to-mark ratio is 1:1 again.

## Adding uncertainty

```{r}
ggplot(ph_summary, aes(arm, mean_ph)) +
  geom_pointrange(aes(ymin = mean_ph - sd_ph, ymax = mean_ph + sd_ph))
```

`geom_pointrange()` shows the mean as a point and ±SD as a line. The standard "mean ± SD" mark.

## Layered: raw rows under a summary mark

```{r}
ggplot() +
  geom_quasirandom(data = table_02, aes(arm, ph), alpha = 0.5) +
  geom_pointrange(
    data = ph_summary,
    aes(arm, mean_ph, ymin = mean_ph - sd_ph, ymax = mean_ph + sd_ph),
    color = "red", size = 0.8
  )
```

Two layers. Two data tables. One figure. Layers can have different data.

## `geom_violin`

```{r}
ggplot(table_02, aes(arm, ph)) +
  geom_violin(fill = "lightblue")
```

Shape + summary in one mark. The smoothed sibling of `geom_boxplot` for comparing distributions across groups.

## `geom_density`

```{r}
ggplot(table_02, aes(ph)) +
  geom_density(fill = "lightblue")
```

A smoothed density curve for one continuous variable. The explicit version of the "fat parts" idea from earlier.

## Layered: violin + quasirandom

```{r}
ggplot(table_02, aes(arm, ph)) +
  geom_violin(fill = "lightblue") +
  geom_quasirandom(alpha = 0.6)
```

Group shape from the violin, individual rows from quasirandom on top.

## Faceted plots from long-format data

```{r}
table_02 |>
  pivot_longer(
    cols = c(ph, nugent_score, crp_blood),
    names_to = "measurement",
    values_to = "value"
  ) |>
  ggplot(aes(arm, value)) +
  geom_boxplot() +
  facet_wrap(~ measurement, scales = "free_y")
```

The `pivot_longer` from the wrangling half *enables* this kind of small-multiples plot. Each facet sees one slice of the long table.

## Choosing a distribution geom

- One variable's distribution → `geom_histogram` or `geom_density`
- Compare many groups concisely → `geom_boxplot`
- Show every observation → `geom_quasirandom`
- Shape + observations together → `geom_violin` + `geom_quasirandom`

The choice depends on what you want the reader to see.

## Exercise 2 {.your-turn}

Open `notebooks/04-visualization.qmd` in your Codespace.

<br>

The modes flip: whoever was agentic in Exercise 1 is manual now, and vice versa.

<br>

```{r}
countdown::countdown(30)
```
